The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-07 rational.tex
_

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! trace+ = trace^+
%! trace* = trace^{\times}
%! ^^^ = ^{-1}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! LeftPlus = L^+
%! RightPlus = R^+
%! LeftTimes = L^{\times}
%! RightTimes = R^{\times}

@@

We start with the semantics for {{langRev}} defined in Sec
\ref{sec:pi}, which we will systematically rewrite with unification
and then present a ``reverse'' evaluator written in the same style. We
shall then combine these into the semantics for {{langRevEE}}.

@@

%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
% <iso, v, cc, s>_{fwd} &|-->& {[iso, v', cc, s']}_{fwd}
% & & where iso s v |--> (v', s')
% <c1(;)c2, v, cc, s>_{fwd} &|-->& <c1, v, Fst cc c2, s>_{fwd}
% {[c1, v, Fst cc c2, s]}_{fwd} &|-->& <c2, v, Snd c1 cc, s>_{fwd}
% {[c2, v, Snd c1 cc, s]}_{fwd} &|-->& {[ c1(;)c2, v, cc,s ]}_{fwd}
% <c1(+)c2, v', cc, s>_{fwd} &|-->& <c1, v, LeftPlus cc c2, s[v' <><> left v]>_{fwd}
% {[ c1, v, LeftPlus cc c2, s ]}_{fwd} &|-->& {[c1 (+) c2, left v, cc,s ]}_{fwd}
% <c1(+)c2, v', cc, s>_{fwd} &|-->& <c2, v, RightPlus c1 cc, s[v' <><> right v]>_{fwd}
% {[ c2, v, RightPlus c1 cc,s ]}_{fwd} &|-->& {[c1 (+) c2, right v, cc,s ]}_{fwd}
% <c1(*)c2, v', cc, s>_{fwd} &|-->& <c1, v1, LeftTimes cc c2 v2, s'>_{fwd}
% & & where s' = s[v <><> (v1, v2)]
% {[ c1, v1, LeftTimes cc c2 v2,s ]}_{fwd} &|-->& <c2, v2, RightTimes c1 v1 cc, s>_{fwd}
% {[ c2, v2, RightTimes c1 v1 cc,s ]}_{fwd} &|-->& {[ c1 (*) c2, (v1, v2), cc,s ]}_{fwd}

@@

%subcode{opsem} include main
%! columnStyle = rclr
%! bck = \triangleleft
% {[iso, v, cc, s]}_{bck} &|-->& <iso, v', cc, s'>_{bck}
% & & where iso^{dagger} s v |--> (v', s')
% <c1, v, Fst cc c2, s>_{bck} &|-->& <c1(;)c2, v, cc, s>_{bck}
% <c2, v, Snd c1 cc, s>_{bck} &|-->& {[c1, v, Fst cc c2, s]}_{bck}
% {[ c1(;)c2, v, cc, s ]}_{bck} &|-->& {[c2, v, Snd c1 cc, s]}_{bck}
% <c1, v, LeftPlus cc c2,s >_{bck} &|-->& <c1(+)c2,left v, cc>_{bck}
% {[c1 (+) c2, v', cc, s ]}_{bck} &|-->& {[ c1, v, LeftPlus cc c2, s[v' <><> left v] ]}_{bck}
% <c2, v, RightPlus c1 cc, s>_{bck} &|-->& <c1(+)c2, right v, cc>_{bck}
% {[c1 (+) c2, v', cc,s ]}_{bck} &|-->& {[ c2, v, RightPlus c1 cc, s[v' <><> right v] ]}_{bck}
% <c1, v1, LeftTimes cc c2 v2, s>_{bck} &|-->& <c1(*)c2, (v1, v2), cc, s>_{bck}
% <c2, v2, RightTimes c1 v1 cc, s>_{bck} &|-->& {[ c1, v1, LeftTimes cc c2 v2, s ]}_{bck}
% {[ c1 (*) c2, v, cc, s ]}_{bck} &|-->& {[ c2, v2, RightTimes c1 v1 cc, S' ]}_{bck}
% && where s' = s[v <><> (v1, v2)]

\subsection{Semantics of {{langRevEE}} }

@@

%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eps+, v, cc, s>_{fwd} |--> <eps+, left (-v'), cc, s[v <><> right v]>_{bck}
% <eps+, v, cc, s>_{fwd} |--> <eps+, right v', cc, s[v <><> left (-v')]>_{bck}

    \item
%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eta+, v, cc, s>_{bck} |--> <eta+, left (-v'), cc, s[v <><> right v]>_{fwd}
% <eta+, v, cc, s>_{bck} |--> <eta+, right v', cc, s[v <><> left (-v')]>_{fwd}

>>>>

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! trace+ = trace^+
%! trace* = trace^{\times}
%! ^^^ = ^{-1}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! LeftP = L^+
%! RightP = R^+
%! LeftT = L^{\times}
%! RightT = R^{\times}

@@

In this section we develop the semantics for {{langRev}} extended with
negative and fractional types, {{langRevEE}}.  We start with the
semantics for {{langRev}} defined in Sec \ref{sec:pi}, which we will
systematically rewrite until we have the desired {{langRevEE}}
sematics. Broadly this construction takes two steps

\begin{enumerate}
\item We rewrite the semantics in Sec \ref{sec:pi} using unification
  of the values instead of direct pattern matching. This gives us the
  neccessary infrastructure for fractional types, as we will see.

\item We write a reverse interpreter by reversing the direction of the
  `{{|-->}}' reductions. This gives us the neccessary infrastructure for
  negative types.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Unification and Logic Programming}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Extending {{langRev}} Semantics}

@@

%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
% <iso, v, C, s>_{fwd} &|-->& {[iso, v', C, s']}_{fwd}
% & & where iso s v |--> (v', s')
% <c1(;)c2, v, C, s>_{fwd} &|-->& <c1, v, Fst C c2, s>_{fwd}
% {[c1, v, Fst C c2, s]}_{fwd} &|-->& <c2, v, Snd c1 C, s>_{fwd}
% {[c2, v, Snd c1 C, s]}_{fwd} &|-->& {[ c1(;)c2, v, C,s ]}_{fwd}
% <c1(+)c2, v', C, s>_{fwd} &|-->& <c1, v, LeftP C c2, s[v' <><> left v]>_{fwd}
% {[ c1, v, LeftP C c2, s ]}_{fwd} &|-->& {[c1 (+) c2, left v, C,s ]}_{fwd}
% <c1(+)c2, v', C, s>_{fwd} &|-->& <c2, v, RightP c1 C, s[v' <><> right v]>_{fwd}
% {[ c2, v, RightP c1 C,s ]}_{fwd} &|-->& {[c1 (+) c2, right v, C,s ]}_{fwd}
% <c1(*)c2, v', C, s>_{fwd} &|-->& <c1, v1, LeftT C c2 v2, s'>_{fwd}
% & & where s' = s[v <><> (v1, v2)]
% {[ c1, v1, LeftT C c2 v2,s ]}_{fwd} &|-->& <c2, v2, RightT c1 v1 C, s>_{fwd}
% {[ c2, v2, RightT c1 v1 C,s ]}_{fwd} &|-->& {[ c1 (*) c2, (v1, v2), C,s ]}_{fwd}

@@

%subcode{opsem} include main
%! columnStyle = rclr
%! bck = \triangleleft
% {[iso, v, C, s]}_{bck} &|-->& <iso, v', C, s'>_{bck}
% & & where iso^{dagger} s v |--> (v', s')
% <c1, v, Fst C c2, s>_{bck} &|-->& <c1(;)c2, v, C, s>_{bck}
% <c2, v, Snd c1 C, s>_{bck} &|-->& {[c1, v, Fst C c2, s]}_{bck}
% {[ c1(;)c2, v, C, s ]}_{bck} &|-->& {[c2, v, Snd c1 C, s]}_{bck}
% <c1, v, LeftP C c2,s >_{bck} &|-->& <c1(+)c2,left v, C>_{bck}
% {[c1 (+) c2, v', C, s ]}_{bck} &|-->& {[ c1, v, LeftP C c2, s[v' <><> left v] ]}_{bck}
% <c2, v, RightP c1 C, s>_{bck} &|-->& <c1(+)c2, right v, C>_{bck}
% {[c1 (+) c2, v', C,s ]}_{bck} &|-->& {[ c2, v, RightP c1 C, s[v' <><> right v] ]}_{bck}
% <c1, v1, LeftT C c2 v2, s>_{bck} &|-->& <c1(*)c2, (v1, v2), C, s>_{bck}
% <c2, v2, RightT c1 v1 C, s>_{bck} &|-->& {[ c1, v1, LeftT C c2 v2, s ]}_{bck}
% {[ c1 (*) c2, v, C, s ]}_{bck} &|-->& {[ c2, v2, RightT c1 v1 C, S' ]}_{bck}
% && where s' = s[v <><> (v1, v2)]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Semantics of {{langRevEE}} }

The previous section has defined all the pieces needed to define the
semantics of {{langRevEE}} and we piece them together here.

@@

%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eps+, v, C, s>_{fwd} |--> <eps+, left (-v'), C, s[v <><> right v]>_{bck}
% <eps+, v, C, s>_{fwd} |--> <eps+, right v', C, s[v <><> left (-v')]>_{bck}

    \item
%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% <eta+, v, C, s>_{bck} |--> <eta+, left (-v'), C, s[v <><> right v]>_{fwd}
% <eta+, v, C, s>_{bck} |--> <eta+, right v', C, s[v <><> left (-v')]>_{fwd}
